Towards an Automated Sportscasting System for Chess

### 2011-05-18 paper/paper.tex
Updated ClassifyLabeledData to identify false positives/negatives; + paper tex and pdf

\title{Towards an Automated Sportscasting System for Chess}
\author{
  Charles Chen \\
  {\tt charleschen@berkeley.edu} \\
\And
  Richard Shin \\
  {\tt ricshin@berkeley.edu} \\
\And
  Jinghao Yan \\
  {\tt jinghao@berkeley.edu} \\
}
\date{}

@@

\begin{abstract}
We investigate applications of statistical learning to automated sportscasting for chess. In particular, this project studies methods to identify high level structures in chess games using support vector machines and autoencoders. 
\end{abstract}

@@

\paragraph{Automated sportscasting.} Chen and Mooney~\shortcite{chen} present a commentator system that learns language from sportscasts of simulated RoboCup games. Their system, which is trained on ambiguous supervision in the form of textual human commentaries and simulation states of the soccer games, simultaneously tries to establish correspondences between the commentaries and the simulation states. Their work focuses on generating meaningful English commentary from relatively simple topics such as ‘Player 1 kicked the ball to Player 2’, while our work focuses more on identifying complex topics.

@@

\begin{table*}
\centering
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Name} & \textbf{Description} \\ \hline
\tt{attackPossible} & The current player has opportunity to attack. \\
\tt{backrankWeakness} & There exists a future checkmate by a rook or queen on the edges of the board. \\
\tt{bishopPair} & There exists a decisive strategic advantage for the player with two bishops. \\
\tt{clearance} & There exists a maneuver to break the pawn structure. \\ 
\tt{closure} & There exists a maneuver to fix the pawn structure. \\
\tt{decoy} & There exists a maneuver to distract a crucial piece. \\
\tt{defensePressure} & There exists a maneuver to break the opposing King’s defense. \\
\tt{deflection} & There exists a maneuver to deflect an attack.  \\ 
\tt{discoveredAttack} & There exists a maneuver to allow for a discovered attack (an attack that was blocked by a friendly piece.) \\
\tt{doubleAttack} & There exists a move that simultaneously attacks at least two crucial squares. \\
\tt{drawCombinationsStalemate} & There exists a combination that allows for decisive material advantage or immediate checkmate. \\
\tt{enclosedKings} & There exists a maneuver to trap the opposing King. \\
\tt{freeingBlockingSquares} & There exists a move to free a crucial square. \\
\tt{goodEndgameChances} & There exists a move to allow for a favorable endgame. \\
\tt{inBetweenMove} & There exists an interrupting move that allows for decisive material advantage of immediate checkmate. \\ 
\tt{interference} & There exists a maneuver to block a crucial piece. \\ 
\tt{interruption} & There exists a maneuver that interrupts the opponent’s tactics. \\
\tt{kingInCenter} & There exists a decisive strategic advantage with the player with the King in the center of the board. \\
\tt{kingToBadSquares} & There exists a combination that forces the opposing King to move to a bad square. \\
\tt{longDiagonals} & There exists a strategic advantage to the player that controls squares on the long diagonal. \\
\tt{matingAttack} & There exists a decisive mating attack. \\
\tt{openingClosingDiagonals} & There exists a maneuver to clear a diagonal. \\
\tt{openingClosingFiles} & There exists a maneuver to clear a file. \\
\tt{overload} & There exists a combination that makes a defending piece defend too many squares at once. \\
\tt{passedPawn} & There exists a combination that allows for a passed pawn (a pawn with no opposing pawns in front of it) \\
\tt{pawnPromotion} & There exists a combination that allows for a pawn promotion. \\
\tt{perpetualCheck} & There exists a combination that forces a perpetual check/draw. \\
\tt{pin} & There exists a combination that allows for a pin. \\
\tt{pullingPiecesToBadSquares} & There exists a combination that forces an opposing piece to move to a bad square. \\
\tt{queenCapture} & There exists a combination to capture the opposing Queen. \\
\tt{quietMates} & There exists a quiet move (one that looks inconsequential on the surface) leading to mate. \\
\tt{removalOfDefence} & There exists a maneuver to remove an opponent’s defender. \\
\tt{unpinning} & There exists a maneuver to unpin a piece. \\
\tt{wrongBishop} & The opposing player’s only bishop is limited in use because of the pawn structure of the game. \\ \hline
\end{tabular}
\caption{List of high-level themes we attempt to automatically identify, along with explanations}
\label{table:themes}
\end{table*}

@@

Since we obtained the games from many sources, there was significant overlap in games (approximately 50% of the unlabeled games collected were redundant). Therefore, we also went through all of the downloaded games and eliminated duplicate games: multiple games that had the same initial position and move sequence were considered redundant and only one of each were kept.

@@

\subsection{Preparing the Data}
After parsing each game, we filtered out the tagged games that had more than 10 plies because it will be very difficult to identify the tags that apply to a complex sequence of moves. Among the games that we kept, we also kept only the first few positions and moves (the ``prefix'' of each game, as described earlier), a parameter that can be adjusted; for our experiments, we kept just the initial board specified in each game. We then generate labels for each game based on the information from the source websites. For example, games in a file that corresponds to a lesson called ``Double Attack'' are labeled doubleAttack. In addition, some of those games may appear in other lessons, and consequently will have multiple labels. We also segment the games further by the color of the player that makes the first move in the game, in addition to the tags of the game in order to account for the white-bias in the training data.

@@

Other strategies for unsupervised feature learning include restricted Boltzmann machines (which are undirected), deep belief nets (stacked RBMs, akin to stacked auto-encoders), and denoising autoencoders (which try to reconstruct some input from a corrupted version of it). We only evaluated sparse stacked auto-encoders in this paper.

@@

As a point of comparison to auto-encoders, we extracted chess-specific features manually from a given chess position (configuration of pieces on the board) and the move that resulted in that position; Table~\ref{table:features} contains a list. The position-based features include information such as where the pieces are on the board and potential moves, while the move-based features involve binary information such as if the move that resulted in this position was a capturing move or resulted in a check.

\section{Experimental methodology}
We processed all of the labeled training data to obtain a mapping from each unique game prefix (for most of the experiments, just the initial specified configuration of the board in the labeled game) to the labels that apply to it; most of the prefixes only had one label, while others had two or three. We then segmented the data into three disjoint sets: 90% for training, 5% as the validation set (for use with tuning our classifiers), and 5% as the test set. The segmenting was done by random selection from the set of all labeled games.

@@

\subsection{Classifying with the auto-encoder}
There exists several methods for using the auto-encoder for the task at hand. First, we can consider using the output of the last hidden layer as the feature vector for input into the SVM classifier, just as we did with the manually-generated features. To do this,

\subsection{Random baseline}
In addition, we compared the predictions made by SVM to those made randomly where each game was assigned a tag at random based on the proportion of the games that have that tag. The random predictions were made 1,000 times, with the predictions made independently at random, and the error rates were averaged.

\section{Experimental results}
\subsection{SVM Based Structured Classifier}
We built a support vector machine for structured classification of games. We filtered our tagged games by the number of plies (number of individual moves), generated mappings between tags and game prefixes (some initial position plus a sequence of moves), segmented our tagged games into three groups (training, validation and test), and called SVMLight with that data, an external library for SVM.

TALK ABOUT THE AUTOENCODER EXPERIMENTS

\subsection{Results}
Since each game has very few tags, most of the binary classifications are negative. As a result, the model is heavily biased toward negative results. Our results bear this disposition:

INSERT TABLE HERE

The basic classifier generates classifications with relatively high precision compared to a random classifier. However, the recall is much lower than that of the random classifier. As a result, the F1 score (a combination of the precision and recall scores) of the SVM-based model is no better than that of the random model.

We then extended the tagging to also incorporate information about the player’s color because there is a bias. For the predictions from the SVM model, this modification improved recall at a minor cost to precision, resulting in a higher F1 score. In contrast, the random predictions got much worse: Both the precision and recall scores fell as a result of this change, which adds further bias and reduces the amount of data per tag (since games are now spread over twice as many tags).

When we delved deeper to analyze the sources of errors and true positives, we arrived at a surprising result. We found that deflection and removalOfDefence are much more likely to cause false negatives than other tags (Table \ref{table:negatives}). In addition, we found that all of the true positives are for the quietMates tag.

\begin{table}
\begin{tabular}{lr}
\hline
\textbf{Tag} & \textbf{Error rate} \\ \hline
\tt{queenCapture}  & 0.006452 \\
\tt{kingInCenter}  & 0.006452 \\
\tt{perpetualCheck}  & 0.006452 \\
\tt{drawCombinationsStalemate}  & 0.038710 \\
\tt{bishopPair}  & 0.006452 \\
\tt{doubleAttack}  & 0.006452 \\
\tt{clearance}  & 0.025806 \\
\tt{openingClosingDiagonals}  & 0.006452 \\
\tt{openingClosingFiles}  & 0.006452 \\
\tt{enclosedKings}  & 0.012903 \\
\tt{discoveredAttack}  & 0.025806 \\
\tt{passedPawn}  & 0.012903 \\
\tt{quietMates}  & 0.038710 \\
\tt{pin}  & 0.051613 \\
\tt{removalOfDefence}  & 0.225806 \\
\tt{deflection}  & 0.251613 \\
\tt{decoy}  & 0.096774 \\
\tt{interference}  & 0.038710 \\
\tt{backrankWeakness}  & 0.032258 \\
\tt{freeingBlockingSquares}  & 0.019355 \\
\tt{pawnPromotion}  & 0.064516 \\
\tt{pullingPiecesToBadSquares}  & 0.006452
\end{tabular} 
\caption{False negative and false positive rates for various tags.}
\label{table:negatives}
\end{table}

Upon closer examination, it was clear that the “checkmate move” feature contributed to that: Any board that is in a checkmate situation is likely to be a quietMate. This interesting result demonstrates the limitations of pursuing a general approach for a highly-complex and specific game like chess.

We sought to improve upon this baseline with an autoencoder. TALK ABOUT AUTOENCODER RESULTS

\section{Implementation} % Parallelization and Scaling
Running serially, the autoencoder takes a long time to complete, even on simple inputs. Therefore, we examined ways to parallelize the task in order to fully leverage the capacity of our 8-core server with hyper-threading (16 virtual cores). We implemented a local version of MapReduce that creates 16 execution threads.

\subsection{Discussion}
We discovered that there is an inherent trade off between how general the approach is and the quality of the outcomes. In pursuing this project, we attempted to maintain an approach that is as general as possible so that it can be applied to other contexts. However, in attempting to maintain generality, we sacrificed access to powerful but chess-specific tools like game engines, resulting in a classifier that is more surface-level (yet slower) than existing chess analysis tools that leverage the game engine.

\section{Future Work}
There are three areas in which future work can make progress on automated sportscasting. It could improve the performance of the autoencoder so that it can be trained on a larger data set. Furthermore, it could focus on trying different methods to improve the classification quality by leveraging chess-specific tools and features like chess engines and evaluation tools. Moreover, it could investigate an orthogonal path by applying our general approach to other games, including complex ones like StarCraft and simpler ones like Checkers.

\subsection{Performance}
One of the main limitations of our work is performance. Due to how slowly the autoencoder performs, we were not able to fully exploit the millions of games that we have available to generate features for our unsupervised system. Therefore, extensions to our work should investigate ways to improve the performance so that the autoencoder can process more inputs and thus generate more relevant feature functions.

\paragraph{Distributed MapReduce.}
Our autoencoder takes advantage of some of the parallel processing capacity of the host system through a local MapReduce scheme. Extending that with a distributed MapReduce framework like Hadoop may improve performance if the additional cost of network I/O are outweighed by the benefits of parallelization.

\paragraph{Using the GPU.}
Another option that we partially explored is the exploitation of the processing power of the graphics processing unit (GPU). Optimization is a computation-heavy task, and the GPU is specifically designed for massively-parallel floating point computations. We investigated using JavaCL to parallelize the autoencoder, and demonstrated that it is definitely superior at parallelizable and computation-intensive tasks like matrix multiplication, because the benefits of greater throughput through greater parallelization outweigh the costs of memory transfer.

\subsection{Classification Quality}
Future work should emphasize the classification quality. We demonstrated that general approaches may not be effective enough to be useful for complex games like chess. Therefore, future projects should seek improvements by better taking advantage of the chess-specific tools and libraries that are available like game engines and weight functions. If the classifier incorporated chess-specific features like those generated by chess engines, it could be much better at identifying the types of strategies that are available to the player. For example, a deep traversal of many future paths by a chess engine may reveal that an attack is possible (attackPossible) or that the player can force a pawn promotion (pawnPromotion). While a purely computational approach to identifying opportunities and strategies may provide higher-quality or more reliable assessments, a statistical approach may be more robust. %talk about what i mean by robust

One potential improvement that does not involve increased reliance on chess-specific technologies is to implement hierarchical tags so that matingAttack, doubleAttack, and similar tags imply a more general tag like attackPossible. By doing so, there will be a greater balance of positives and negatives, and labels that applied to games but were previously considered wrong are now considered correct. By incorporating hierarchical structure to tags, the system can jointly train and classify games.

Another possible improvement is to use a different classifier than SVM, such as the perceptron or a maximum entropy classifier. One issue with the SVM is that often the tags are not so clear-cut: what does it mean that an endgame has “good chances”? Rather, there are degrees of how good an endgame is to the player. Therefore, a softer approach like maximum entropy may produce better results.

@@

\section{Conclusion}

\section{Acknowledgements}
We would like to thank Dan Klein for teaching us statistical machine learning techniques.
\section*{Appendix}

>>>>

\title{Towards an Automated Sportscasting System for Chess}
\author{
  Charles Chen \\
	UC Berkeley \\
	Berkeley, CA 94720 \\
  {\tt \small{charleschen@berkeley.edu}} \\
\And
  Richard Shin \\
	UC Berkeley \\
	Berkeley, CA 94720 \\
  {\tt \small{ricshin@berkeley.edu}} \\
\And
  Jinghao Yan \\
	UC Berkeley \\
	Berkeley, CA 94720 \\
  {\tt \small{jinghao@berkeley.edu}} \\
}

@@

\begin{abstract}
We investigate applications of statistical learning to automated sportscasting for chess. In particular, this project studies methods to automatically identify high level structures in chess games using support vector machines and auto-encoders. We then compare the results of the two different approaches and showed improvements over the SVM model using a stacked auto-encoder. Finally, we conclude with results, analysis of errors, and ideas for future work.
\end{abstract}

@@

\paragraph{Automated sportscasting.} Chen and Mooney~\shortcite{chen} present a commentator system that learns language from sportscasts of simulated RoboCup games. Their system, which is trained on ambiguous supervision in the form of textual human commentaries and simulation states of the soccer games, simultaneously tries to establish correspondences between the commentaries and the simulation states. Their work focuses on generating meaningful English commentary from relatively simple topics such as ``Player 1 kicked the ball to Player 2'', while our work focuses less on text generation and more on identifying complex topics.

@@

\begin{table*}
\centering
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Name} & \textbf{Description} \\ \hline
\tt{attackPossible} & The current player has opportunity to attack. \\
\tt{backrankWeakness} & There exists a future checkmate by a rook or queen on the edges of the board. \\
\tt{bishopPair} & There exists a decisive strategic advantage for the player with two bishops. \\
\tt{clearance} & There exists a maneuver to break the pawn structure. \\ 
\tt{closure} & There exists a maneuver to fix the pawn structure. \\
\tt{decoy} & There exists a maneuver to distract a crucial piece. \\
\tt{defensePressure} & There exists a maneuver to break the opposing King’s defense. \\
\tt{deflection} & There exists a maneuver to deflect an attack.  \\ 
\tt{discoveredAttack} & There exists a maneuver to allow for a discovered attack (an attack that was blocked by a friendly piece.) \\
\tt{doubleAttack} & There exists a move that simultaneously attacks at least two crucial squares. \\
\tt{drawCombinationsStalemate} & There exists a combination that allows for decisive material advantage or immediate checkmate. \\
\tt{enclosedKings} & There exists a maneuver to trap the opposing King. \\
\tt{freeingBlockingSquares} & There exists a move to free a crucial square. \\
\tt{goodEndgameChances} & There exists a move to allow for a favorable endgame. \\
\tt{inBetweenMove} & There exists an interrupting move that allows for decisive material advantage of immediate checkmate. \\ 
\tt{interference} & There exists a maneuver to block a crucial piece. \\ 
\tt{interruption} & There exists a maneuver that interrupts the opponent’s tactics. \\
\tt{kingInCenter} & There exists a decisive strategic advantage for the player with the King in the center of the board. \\
\tt{kingToBadSquares} & There exists a combination that forces the opposing King to move to a bad square. \\
\tt{longDiagonals} & There exists a strategic advantage for the player that controls squares on the long diagonal. \\
\tt{matingAttack} & There exists a decisive mating attack. \\
\tt{openingClosingDiagonals} & There exists a maneuver to clear a diagonal. \\
\tt{openingClosingFiles} & There exists a maneuver to clear a file. \\
\tt{overload} & There exists a combination that makes a defending piece defend too many squares at once. \\
\tt{passedPawn} & There exists a combination that allows for a passed pawn (a pawn with no opposing pawns in front of it.) \\
\tt{pawnPromotion} & There exists a combination that allows for a pawn promotion. \\
\tt{perpetualCheck} & There exists a combination that forces a perpetual check/draw. \\
\tt{pin} & There exists a combination that allows for a pin. \\
\tt{pullingPiecesToBadSquares} & There exists a combination that forces an opposing piece to move to a bad square. \\
\tt{queenCapture} & There exists a combination to capture the opposing Queen. \\
\tt{quietMates} & There exists a quiet move (one that looks inconsequential on the surface) leading to mate. \\
\tt{removalOfDefence} & There exists a maneuver to remove an opponent’s defender. \\
\tt{unpinning} & There exists a maneuver to unpin a piece. \\
\tt{wrongBishop} & The opponent’s only bishop is limited in use because of the pawn structure of the game. \\ \hline
\end{tabular}
\caption{List of high-level themes we attempt to automatically identify, along with explanations}
\label{table:themes}
\end{table*}

@@

Since we obtained the games from many sources, there was significant overlap in games (approximately 50\% of the unlabeled games collected were redundant). Therefore, we also went through all of the downloaded games and eliminated duplicate games: multiple games that had the same initial position and move sequence were considered redundant and only one of each were kept.

@@

\subsection{Preparing the Data}
After parsing each game, we filtered out the tagged games that had more than 10 plies because it will be very difficult to identify the tags that apply to a complex sequence of moves. Among the games that we kept, we also kept only the first few positions and moves (the ``prefix'' of each game, as described earlier), a parameter that can be adjusted; for our experiments, we kept just the initial board specified in each game. We then generate labels for each game based on the information from the source websites. For example, games in a file that corresponds to a lesson called ``Double Attack'' are labeled {\tt doubleAttack}. In addition, some of those games may appear in other lessons, and consequently will have multiple labels. We also segment the games further by the color of the player that makes the first move in the game, in addition to the labels of the game, since most of the labels in Table~\ref{table:themes} only refer to a specific player; for instance, some of the games labeled with {\tt backrankWeakness} may have the white player with the weakness, while others show the black player with the weakness.

@@

Other strategies for unsupervised feature learning include restricted Boltzmann machines (which are undirected), deep belief nets (stacked RBMs, akin to stacked auto-encoders), and denoising auto-encoders (which try to reconstruct some input from a corrupted version of it). We only evaluated sparse stacked auto-encoders in this paper.

The auto-encoder requires training data in the form of vectors where all the elements are between 0 and 1, or the range of the sigmoid function. We generated the training vectors by, for some chess board and for every type of piece (knight, rook, etc.), whether a white one (valued 1.0), a black one (valued 0.0), or none at all (valued 0.5) was present in every square of the chessboard.

@@

As a point of comparison to auto-encoders, we extracted chess-specific features manually from a given chess position (configuration of pieces on the board) and the move that resulted in that position; Table~\ref{table:features} contains a list. The position-based features include information such as where the pieces are on the board and potential moves, while the move-based features involve binary information such as if the move that resulted in this position was a capturing move or resulted in a check.  The resulting feature vector was then L2-normalized to fit the specifications of the support vector machine.

\section{Experimental methodology}
We processed all of the labeled training data to obtain a mapping from each unique game prefix (for most of the experiments, just the initial specified configuration of the board in the labeled game) to the labels that apply to it; most of the prefixes only had one label, while others had two or three. We then segmented the data into three disjoint sets: 90\% for training, 5\% as the validation set (for use with tuning our classifiers), and 5\% as the test set. The segmenting was done by random selection from the set of all labeled games.

@@

\subsection{Classifying with the auto-encoder}
\label{sec:classifying-auto-encoder}
There exists several methods for using the auto-encoder for the task at hand. First, we can consider using the output of the last hidden layer as the feature vector for input into the SVM classifier, just as we did with the manually-generated features. To do this, we first perform unsupervised fine-tuning; we combine all the greedily-trained auto-encoders to make a many-layered neural network intended to reconstruct the input layer in its output layer, after which we again minimize the sum-squared error between the input and the output layers. We can then use the output of the hidden layer (i.e., not the one that reconstructs the input, but the middle layer which provides the most high-level representation of the input) as the feature vector.

We can also use semi-supervised learning methods, in which after greedily training the auto-encoders, we combine all the hidden layers of the auto-encoders and add an output layer at the end. For every labeled input game, we construct input-output vectors such that the output encodes the gold labels of the input. We then train the resulting multi-layer neural network to minimize the sum-squared error; this acts similarly to a multiclass logistic regression, except with a different objective at the last layer.

To train these neural networks, we used batch learning over all the examples at once. The loss function contained components for the sum-squared error as well as an L2 regularization term and a sparsity penalty based on the Kullback-Leibler divergence between the desired average activation of the hidden nodes and the actual average activation. We used backpropagation to compute the gradient of the loss function and L-BFGS to optimize the parameters as to minimize the loss function~\cite{ng-lecturenotes}. Running serially, the training takes a long time to complete, even on simple inputs. Therefore, we examined ways to parallelize the task in order to fully leverage the capacity of modern many-core machines; we achieved a significant speedup by parallelizing the loss function evaluation to 16 execution threads.

\subsection{Random baseline}
To use as a baseline and ensure that our classifiers are giving sensible results, we randomly made predictions for every test game, where each one assigned each possible label at random, with the possibility of the label being given based on the proportion of the training games which have that label. The random predictions were made 1,000 times, with the predictions made independently at random, and the error rates were averaged.

\section{Experimental results}
\subsection{SVM classifier}
\begin{table*}
\centering
\begin{tabular}{lccccccc}
\hline
\textbf{Type} & \textbf{TP} & \textbf{FP} & \textbf{TN} & \textbf{FN} & \textbf{Precision} & \textbf{Recall} & \textbf{F1} \\ \hline
Unsegmented & 0.21\% & 0.00\% & 96.89\% & 2.90\% & 100\% & 6.71\% & 12.57\% \\
Baseline (Random) & 0.42\% & 2.76\% & 94.12\% & 2.69\% & 13.32\% & 13.65\% & 13.48\% \\
\hline
Segmented & 0.13\% & 0.02\% & 98.34\% & 1.51\% & 86.67\% & 7.65\% & 14.05\% \\
Baseline (Random) & 0.12\% & 1.49\% & 96.87\% & 1.51\% & 7.65\% & 7.55\% & 7.60\% \\
\end{tabular} 
\caption{Classification results for the games that have not been segmented by the color of the active player, and those that have been. In the former, precision clearly outperforms a random baseline, but recall falls short, resulting in a lower F1 score. In the latter, precision has fallen relative to the unsegmented version, but recall has risen, resulting in a slightly higher F1 score. In contrast, the random model became much worse in both ways.}
\label{table:svmResults}
\end{table*}

We first describe the results achieved when we trained using the manually-specified feature extractors. Since each game has very few labels, most of the binary classification results are negative. Therefore, the model is heavily biased toward negative results, as shown in Table~\ref{table:svmResults}.

This classification method had relatively high precision compared to the random baseline. However, the recall is much lower than that of the random classifier. As a result, the F1 score (a combination of the precision and recall scores) of the SVM-based model is no better than that of the random model.

When we extended the labeling to also incorporate the first mover's color information, the SVM model's recall was improved at a minor cost to precision, resulting in a higher F1 score. In contrast, the random predictions became much worse: Both the precision and recall scores fell as a result of this change, which reduced the evenness of the label distribution among the training games, and reduced the amount of games with each label (since games are now spread over twice as many tags).
\begin{table}
\centering
\begin{tabular}{lr}
\hline
\textbf{Tag} & \textbf{Error rate} \\ \hline
\tt{queenCapture}  & 0.6452\% \\
\tt{kingInCenter}  & 0.6452\% \\
\tt{perpetualCheck}  & 0.6452\% \\
\tt{drawCombinationsStalemate}  & 3.8710\% \\
\tt{bishopPair}  & 0.6452\% \\
\tt{doubleAttack}  & 0.6452\% \\
\tt{clearance}  & 2.5806\% \\
\tt{openingClosingDiagonals}  & 0.6452\% \\
\tt{openingClosingFiles}  & 0.6452\% \\
\tt{enclosedKings}  & 1.2903\% \\
\tt{discoveredAttack}  & 2.5806\% \\
\tt{passedPawn}  & 1.2903\% \\
\tt{quietMates}  & 3.8710\% \\
\tt{pin}  & 5.1613\% \\
\tt{removalOfDefence}  & 22.5806\% \\
\tt{deflection}  & 25.1613\% \\
\tt{decoy}  & 9.6774\% \\
\tt{interference}  & 3.8710\% \\
\tt{backrankWeakness}  & 3.2258\% \\
\tt{freeingBlockingSquares}  & 1.9355\% \\
\tt{pawnPromotion}  & 6.4516\% \\
\tt{pullingPiecesToBadSquares}  & 0.6452\%
\end{tabular} 
\caption{Percent of incorrect classification (false negatives and false positives) for various tags.}
\label{table:negatives}
\end{table}

Given the high precision but low recall, we delved deeper to analyze the sources of errors and true positives. As we show in Table~\ref{table:negatives}, we found that {\tt deflection} and {\tt removalOfDefence} are much more likely to cause false negatives than other tags. In addition, we found that all of the true positives are for the {\tt quietMates} tag. 9 of the 11 true positive {\tt quietMates} games were classified correctly using only the features that describe the contents of the board. On the other hand, 5 true positive {\\tt quietMates} games were correctly classified when using only features about the possible moves in that position. Interestingly, disabling the ``checkmate move'' feature did not affect the number of true positives.

\subsection{Classifying with the auto-encoder}
\begin{table*}
\centering
\begin{tabular}{lcccccccc}
\hline
\textbf{Type} & \textbf{Hidden layers} & \textbf{TP} & \textbf{FP} & \textbf{TN} & \textbf{FN} & \textbf{Precision} & \textbf{Recall} & \textbf{F1} \\ \hline
Unsegmented & 0 & 1.49\% & 3.62\% & 93.26\% & 1.61\% & 29.26\% & 48.17\% & 36.40\% \\
Unsegmented & 1 & 1.35\% & 3.83\% & 93.06\% & 1.76\% & 26.01\% & 43.29\% & 32.49\% \\
Unsegmented & 2 & 1.40\% & 4.48\% & 92.41\% & 1.71\% & 23.87\% & 45.12\% & 31.22\% \\
Unsegmented & 3 & 1.40\% & 4.48\% & 92.41\% & 1.71\% & 23.87\% & 45.12\% & 31.22\% \\
\hline
Segmented & 0 & 0.38\% & 1.01\% & 97.35\% & 1.26\% & 27.08\% & 22.94\% & 24.84\% \\
Segmented & 1 & 0.49\% & 2.90\% & 95.46\% & 1.15\% & 14.49\% & 30.00\% & 19.54\% \\
Segmented & 2 & 0.22\% & 1.27\% & 97.09\% & 1.42\% & 14.84\% & 13.53\% & 14.15\% \\
Segmented & 3 & 0.59\% & 3.89\% & 94.47\% & 1.05\% & 13.12\% & 35.88\% & 19.21\% \\
\end{tabular} 
\caption{Auto-encoder classification results for the games that have not been segmented by the color of the active player, and those that have been, as the number of hidden layers varies.}
\label{table:auto-encoder-results}
\end{table*}

We first tried to use the unsupervised fine-tuned stacked auto-encoder's hidden outputs as the input to the SVM classifiers, as with the manually-generated feature vectors. However, we found that the resulting feature vectors were not very useful for SVM, as the classifiers did not predict any labels as true for all training or test input games.

Instead, we used the semi-supervised classification method as described in Section~\ref{sec:classifying-auto-encoder}. First, we greedily train each hidden layer as an auto-encoder on the training, test, and validation sets. We trained the combined neural network composed of greedily-trained auto-encoders to, for each input, predict an output vector corresponding to the labels that apply to the input---0 if a label did not apply, and 1 if it did. From the output vectors computed with the elements in the validation set, we tuned the threshold we use in considering whether a label is present or not present to maximize the F1 score. In Table~\ref{table:auto-encoder-results}, we report the performance achieved on the test set when using the threshold tuned to maximize the F1 score on the validation set.

These results show higher F1 score than the SVM classifier, with significantly lower precision but also much higher recall. Unfortunately, the results indicate that the representation of the board learned by the auto-encoder does not correlates very well with the concepts we wanted to identify, given that having no hidden layers resulted in higher scores. Adding extra hidden layers seems to not have helped improve the classification results, except in the last instance, where an extra layer improved the F1 score to original levels. It is hard to tell whether some slight change in the auto-encoder training procedure will be able to drastically improve the results, or whether significant changes in the architecture will be required. We also were not able to use the unlabeled chess games we collected with the auto-encoder, due to the long training time required and our relatively slow implementation of its training.

\subsection{Discussion}
In pursuing this project, we attempted to maintain an approach that is as general as possible so that it can be applied to other contexts, and to minimize the amount of engineering required very specific to chess. However, in attempting to maintain generality, we sacrificed access to powerful but chess-specific tools like chess engines.%, resulting in a classifier that is more surface-level (yet slower) than existing chess analysis tools that leverage the game engine.

A chess-specific approach would be able to perfectly identify the more precisely-described tags like {\tt doubleAttack} and {\tt queenCapture} by analyzing the future branches and looking for specific outcomes. In addition, it would be able to identify less concrete tags like {\tt clearance} and {\tt kingToBadSquares}. However, one advantage of the statistical approach is that feature functions can be written once and reused for all inputs. In contrast, in order to effectively identify tags using a chess engine, each tag must have a specific heuristic function written for that express purpose to identify certain properties in future states. A better automated sportscasting for chess will leverage the strengths of each approach by using the chess-specific features when they are available, and combining that with results from statistical learning. For instance, while the chess engine can easily identify imminent mate combinations, it would require a coordination of both the engine and the statistical learning tool to identify threshold functions like {\tt kingToBadSquares}. Machine learning would be used to optimize weights and thresholds (e.g. for whether a king’s position is bad) while the engine would be used to explore future paths.

Based on the performance of our general approach, we believe that there is an inherent trade off between how general the approach is and the quality of the outcomes.

\section{Future Work}
There are three areas in which future work can make progress on automated sportscasting. It could improve the performance of the auto-encoder so that it can be trained on a larger data set. Furthermore, it could focus on trying different methods to improve the classification quality by leveraging chess-specific tools and features like chess engines and evaluation tools. Moreover, it could investigate an orthogonal path by applying our general approach to other games, including complex ones like StarCraft and simpler ones like Checkers.

\subsection{Performance}
One of the main limitations of our work is performance. Due to how slowly the auto-encoder performs, we were not able to fully exploit the millions of games that we have available to generate features for our unsupervised system. Therefore, extensions to our work should investigate ways to improve the performance so that the auto-encoder can process more inputs and thus generate more relevant feature functions.

\paragraph{Distributed MapReduce.}
Our auto-encoder takes advantage of some of the parallel processing capacity of the host system through a local MapReduce scheme. Extending that with a distributed MapReduce framework like Hadoop may improve performance if the additional cost of network I/O are outweighed by the benefits of parallelization.

\paragraph{Using the GPU.}
Another option that we partially explored is the exploitation of the processing power of the graphics processing unit (GPU). Optimization is a computation-heavy task, and the GPU is specifically designed for massively-parallel floating point computations. We investigated using JavaCL to parallelize the auto-encoder, and demonstrated that it is definitely superior at parallelizable and computation-intensive tasks like matrix multiplication, because the benefits of greater throughput through greater parallelization outweigh the costs of memory transfer.

\subsection{Classification Quality}
Future work should emphasize the classification quality. We demonstrated that general approaches may not be effective enough to be useful for complex games like chess. Therefore, future projects should seek improvements by better taking advantage of the chess-specific tools and libraries that are available like game engines and weight functions. If the classifier incorporated chess-specific features like those generated by chess engines, it could be much better at identifying the types of strategies that are available to the player. For example, a deep traversal of many future paths by a chess engine may reveal that an attack is possible ({\tt attackPossible}) or that the player can force a pawn promotion ({\tt pawnPromotion}). While a purely computational approach to identifying opportunities and strategies may provide higher-quality or more reliable assessments, a statistical approach may be more robust to data sparsity.

One potential improvement that does not involve increased reliance on chess-specific technologies is to implement hierarchical tags so that {\tt matingAttack}, {\tt doubleAttack}, and similar tags imply a more general tag like {\tt attackPossible}. By doing so, there will be a greater balance of positives and negatives, and labels that applied to games but were previously considered wrong are now considered correct. By incorporating hierarchical structure to tags, the system can jointly train and classify games.

Another possible improvement is to use a different classifier than an SVM, such as the perceptron or a maximum entropy classifier. One issue with the SVM is that often the tags are not so clear-cut: what does it mean that an endgame has “good chances”? Rather, there are degrees of how good an endgame is to the player. Therefore, a softer approach like maximum entropy may produce better results.

@@

\section{Conclusion}
We built two statistical learning systems to classify high-level positional concepts for chess, and described various issues involved in implementing these systems. Both classifiers are feature driven and classified by a support vector machine. The first system uses manually generated features for a chessboard configuration and the move that led to it, while the second system uses a sparse auto-encoder to automatically discover high-level structures (and hence features) for a given chessboard configuration in an unsupervised fashion. We found that the auto-encoder-based classifier outperformed the manual feature vector-based classifier, which outperformed the random baseline classifier: on the same segmented data set, the random baseline classifier had an F1 score of $\mathbf{7.60}$, the best manual feature vector-based classifier had an F1 score of $\mathbf{14.05}$, and the best auto-encoder-based classifier had an F1 score of $\mathbf{24.84}$.

\section{Acknowledgements}
We would like to thank Dan Klein for teaching us statistical machine learning techniques, and Soda Hall for accommodating us for many nights.
